Left-linear rules in adhesive categories

### 2024-02-14 icalp/fusione.tex

  \item
    If $1$ is a terminal object in $\X$, the unique
    arrow $X\to 1$ from an object $X$ will be denoted by
    $!_X$. Similarly, if $0$ is initial in $\X$ then $?_X$ will denote
    the unique arrow $0\to
    X$. %When $\X$ is $\Set$ and $1$ is a singleton, $\delta_x$ will denote the arrow $1\to X$ with value $x\in X$.

@@

\begin{example}
\todo{esempi (no idea)}
\end{example}

@@

\begin{example}[E-graphs]
  Consider the category $\cat{EGraphs}$, where objects are (directed)
  graphs endowed with an equivalence over nodes and arrows are graph
  morphisms which preserve the equivalence, as considered
  in~\cite{BaldanGM06}, closely related to e-graphs~\cite{WNW:egg}. 
  Formally, $\cat{EGraphs}$ can be seen as the
  subcategory of the presheaf
  $[E \rightrightarrows V \to Q, \cat{Set}]$ where objects are
  constrained to have the component $Q$ surjective. Explicitly, an
  e-graph $G = \langle s_G, t_G, q_G \rangle$ where
  $s_G, t_G: E_G \rightrightarrows V_G$ provides the graphical
  structure, while the surjective function $q_G : V_G \to Q_G$ maps
  each node to the corresponding equivalence class.

  A morphism in $\cat{EGraph}$ is mono if the components over $E$
  and $V$ are mono, i.e., if it is mono as a morphism in
  $\cat{Graph}$. It is \emph{regular mono} if also the component on
  $Q$ is mono, i.e. if it reflects equivalence classes besides
  preserving them.

  \todo[inline]{$\cat{EGraph}$ is quasi-adhesive and well-switching}  
\end{example}

@@

\begin{restatable}[Globality of independence]{lemma}{lemIndepGlobalLeft}
	\label{lem:indep-global-left}
	Let $\X$ be an $\mathcal{M}$-adhesive category and
	$(\X,\R)$ a strong enforcing rewriting system.
	Consider a derivation
	$\der{D}=\{\dder{D}_i\}_{i=0}^2$ and suppose that $(i_0,i_1)$ is an independence pair between $\dder{D}_0$ and $\dder{D}_1$,
	$(a_0,a_1)$ one between $\dder{D}_1$ and $\dder{D}_2$ and
	$(e_0, e_1)$ one between $\dder{D}_0$ and
	$S_{a_0,a_1}(\dder{D}_2)$. Then the following properties hold 
	\begin{enumerate}
		\item
		\label{lem:indep-global-left:1}
		$S_{e_0,e_1}(\dder{D}_0)$ and $S_{a_0,a_1}(\dder{D}_1)$ are
		switchable; 
		\item
		\label{lem:indep-global-left:2}
		if $(\X, \R)$ is well-switching and $S_{i_0, i_1}(\dder{D}_0)\seqind \dder{D}_2$ with an independence pair $(\alpha_0, \alpha_1)$, then
		$S_{i_0,i_1}(\dder{D}_1)\seqind S_{\alpha_0, \alpha_1}(\dder{D}_2)$.
	\end{enumerate}
\end{restatable}

  %

@@

\begin{restatable}[Confluence of switchings]{lemma}{lemSwitchConfluence}
	\label{lem:switch-confluence}
	Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X,\R)$ a
	left-linear rewriting system.
	% 
	Consider a derivation $\der{D}=\{\dder{D}_i\}_{i=0}^2$ and suppose
	that $(i_0,i_1)$ is a strong independence pair between $\dder{D}_0$
	and $\dder{D}_1$, $(a_0,a_1)$ one between $\dder{D}_1$ and
	$\dder{D}_2$ and $(e_0, e_1)$ one between $\dder{D}_0$ and
	$S_{a_0,a_1}(\dder{D}_2)$.\todo{P: non e' meglio qui limitarci a well-switching, cosi' da evitare riferimenti a strong pairs?}
\end{restatable}

@@

\iffalse
	\begin{example}
		In a generic category $\X$, pushout complements may not exist: in $\Set$ the arrows $?_{2}\colon \emptyset \to 2$ and $!_2\colon 2\to 1$ do not have a pushout complement.

		Moreover, composable arrows $f\colon X\to Y$ and $g\colon Y\to Z$ may have  pushout complements which are non-isomorphic: for instance, in $\Set$ the two squares below are both pushouts.

		\[\xymatrix{2 \ar[r]^{!_2} \ar[d]_{\id{2}}& 1 \ar[d]^{\id{1}} & 2 \ar[r]^{!_2} \ar[d]_{!_2}& 1 \ar[d]^{\id{1}}\\ 2 \ar[r]_{!_2}& 1 & 1 \ar[r]_{\id{1}}& 1}\]
	\end{example}
\fi

@@

\begin{definition} 
	Let $\X$ be an  $\mathcal{M}$-adhesive category, the pair $(\X, \mathcal{M})$ is of \emph{type $\mathbb{B}$} if:
	\begin{enumerate}
		\item every arrow in $\mathcal{M}$ is a coproduct coprojection;
		\item $\X$ has all pushouts;
		\item $\X$ has strict initial objects and, for every object $X$, the arrow $?_X\colon 0\to X $ belongs to $\mathcal{M}$;
		\item all pushouts are $\mathcal{M}$-stable.
	\end{enumerate}
\end{definition}

@@

\begin{proof}   
	Using \Cref{rem:fill} we can build the diagrams in \Cref{fi:first,fi:second,fi:third}.  The first depicts the  switching $\dder{D}_0$ and $\dder{D}_1$ along  $(i_0, i_1)$, the second that of $\dder{D}_1$ and $\dder{D}_2$ along $a_0$ and $a_1$, while the third refers to $\dder{D}_0$ and
	$S_{a_0,a_1}(\dder{D}_2)$.

@@

		\item In addition to the three diagram above, we have a fourth one, depicted in \Cref{fi:fourth}.
		\begin{figure}[h]
			\centering
			$\xymatrix@C=15pt@R=15pt{&&R_0 \ar@/_.8cm/[ddrr]_(.2){\alpha_0}|(.7)\hole \ar[dr]^{\hspace{-1pt}g_1\circ i_0}&& L_2\ar@/^.8cm/[ddll]^(.2){\alpha_1}  \ar[dl]_{m_2\hspace{-2pt}}\\
				&K_0\ar[dr]^{\hspace{-4pt}q_1\circ v_0}\ar@{>->}[dl]_{l_0} \ar@/_.6cm/[ddrr]|(.36)\hole^(.65){\phi_0}\ar[ur]^{r_0}&& G_2 && K_2 \ar@/^.6cm/[ddll]|(.36)\hole_(.65){\phi_2}\ar[dl]_{k_2}\ar@{>->}[lu]_{l_2} \ar[dr]^{r_2}\\
				L_0 \ar@/_.6cm/[ddrr]_(.2){\gamma_0}|(.31)\hole|(.81)\hole \ar[dr]^(.4){\hspace{-4pt}t_0\circ j_0}|(.6)\hole && Q_1 \ar@{>->}[dl]|(.4)\hole_(.65){s_1}\ar[ur]|(.5)\hole^(.7){t_1}&&D_2 \ar[dr]|(.4)\hole^(.65){g_2} \ar@{>->}[ul]|(.5)\hole_(.65){f_2}&&R_2\ar@/^.6cm/[ddll]^(.2){\gamma_1}|(.31)\hole|(.81)\hole\ar[dl]_(.35){h_2}|(.6)\hole\\&H'_1 &&P_4\ar[dr]_{\lambda_1} \ar@{>->}[dl]^{\lambda_0}\ar[ur]^(.4){\zeta_1}\ar@{>->}[ul]_(.4){\zeta_0}&&G_3\\
				L_2 \ar@/^.6cm/[uurr]^(.2){\alpha_1} \ar[ur]_(.35){\hspace{-2pt}s_1\circ
					\alpha_1}|(.61)\hole&&Q_6
				\ar@{>->}[ul]|(.4)\hole_(.6){\xi_0}\ar[dr]|(.5)\hole^(.65){\xi'_0}
				&&Q_7\ar[ur]|(.4)\hole^(.6){\xi'_1}
				\ar@{>->}[dl]|(.5)\hole_(.65){\xi_1} && R_0 \ar[ul]^(.35){g_2\circ
					\alpha_0}|(.61)\hole\ar@/_.6cm/[uull]_(.2){\alpha_0}\\&K_2
				\ar[ur]_{\lambda_0\circ \phi_2} \ar[dr]_{r_2}
				\ar@{>->}[ul]^{l_2}\ar@/^.6cm/[uurr]_(.65){\phi_2}&&H'_2&& K_0
				\ar[ur]_{r_0} \ar[ul]^{\lambda_1\circ \phi_0} \ar@{>->}[dl]^{l_0}
				\ar@/_.6cm/[uull]^(.65){\phi_0}\\&& R_2
				\ar[ur]_{\hspace{-5pt}\xi_1\circ
					\gamma_1}\ar@/^.8cm/[uurr]^(.2){\gamma_1}&& L_0
				\ar[ul]^{\xi'_0\circ \gamma_0\hspace{-7pt}}
				\ar@/_.8cm/[uull]_(.2){\gamma_0} |(.69)\hole 
			}	$
			\caption{Picture for the proof of \Cref{lem:indep-global-left}: switch of $S_{i_0, i_2}(\dder{D}_0)$ and $\dder{D}_2$.}
			\label{fi:fourth}
		\end{figure}

>>>>

  %\item
    %If $1$ is a terminal object in $\X$, the unique
   % arrow $X\to 1$ from an object $X$ will be denoted by
    %$!_X$. Similarly, if $0$ is initial in $\X$ then $?_X$ will denote
    %the unique arrow $0\to X$. %When $\X$ is $\Set$ and $1$ is a singleton, $\delta_x$ will denote the arrow $1\to X$ with value $x\in X$.

@@

%\begin{example}
%\todo{esempi (no idea)}
%\end{example}

@@

\begin{example}[E-graphs]
  Consider the category $\cat{EGraphs}$, where objects are (directed)
  graphs endowed with an equivalence over nodes and arrows are graph
  morphisms which preserve the equivalence, as considered
  in~\cite{BaldanGM06}, closely related to e-graphs~\cite{WNW:egg}. 
  Formally, $\cat{EGraphs}$ can be seen as the
  subcategory of the presheaf
  $[E \rightrightarrows V \to Q, \cat{Set}]$ where objects are
  constrained to have the component $Q$ surjective.

  Explicitly, an
  e-graph $G = \langle s_G, t_G, q_G \rangle$ where
  $s_G, t_G: E_G \rightrightarrows V_G$ provides the graphical
  structure, while the surjective function $q_G : V_G \to Q_G$ maps
  each node to the corresponding equivalence class. Notice that the inclusion functor into  $[E \rightrightarrows V \to Q, \cat{Set}]$ creates pullbacks and pushouts \cite{mac2013categories} so that they are computed componentwise.

  A morphism in $\cat{EGraph}$ is mono if the components over $E$
  and $V$ are mono, i.e., if it is mono as a morphism in
  $\cat{Graph}$. It is regular mono if also the component on
  $Q$ is mono, i.e. if it reflects equivalence classes besides
  preserving them. This characterization of regular monos and the fact that pullbacks and pushouts are computed componenetwise allows us to prove quasiadhesivity of $\cat{EGraphs}$ at once. Moreover, one can deduce that every rewriting system $(\X, \R)$ which is left-linear with respect to $\reg(\cat{EGraphs})$ is strong enforcing: this is done exploiting again the inclusion functor into $[E \rightrightarrows V \to Q, \cat{Set}]$.

  Finally, in applications, one is interested in rewriting systems in which the right half of a rule is always mono but not regular in general, allowing us to identify more nodes. These rewring systems are all well-switching.

  \todo[inline]{Noi prendiamo sistemi di riscrittura left-linear e mono a destra, c'è una refernza?}  
\end{example}

@@

\begin{restatable}[Globality of independence]{lemma}{lemIndepGlobalLeft}
	\label{lem:indep-global-left}
	Let $\X$ be an $\mathcal{M}$-adhesive category and	$(\X,\R)$ a well-switching DPO rewriting system.
For every  derivation
	$\der{D}=\{\dder{D}_i\}_{i=0}^2$  of lenght $3$, the following hold true:
	\begin{enumerate}
		\item
		\label{lem:indep-global-left:1} if $\dder{D}_0\seqind \dder{D}_1$ and there is a switching sequence
		\[\der{D}\shift{(1,2)} \der{D}_1\shift{(0,1)} \der{D}_2\]
		then the last two steps of $\der{D}_2$ are switchable;

		\item
		\label{lem:indep-global-left:2}
		suppose that $\dder{D}_1 \seqind\dder{D}_2$ with independence pair $(\alpha_0, \alpha_1)$ and that, moreover,  $\dder{D}_0\seqind S_{\alpha_0, \alpha_1}(\dder{D}_2)$, if there is a switching sequence
			\[\der{D}\shift{(0,1)} \der{E}_1\shift{(1,2)} \der{E}_2\]
	then the first two steps of $\der{E}_2$ are switchable.
	\end{enumerate}
\end{restatable}
  %

@@

\begin{restatable}[Confluence of switchings]{lemma}{lemSwitchConfluence}
	\label{lem:switch-confluence}
	Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X,\R)$ a	well-switching DPO rewriting system.
	% 
	Consider a derivation $\der{D}=\{\dder{D}_i\}_{i=0}^2$ and suppose that we have the following two switching sequences
	\[\der{D} \shift{(0,1)} \der{D}_1 \shift{(1,2)} \der{D}_2 \shift{(0,1)}\der{D}_3\\ \qquad \der{D}\shift{(1,2)}\der{E}_1\shift{(0,1)} \der{E}_2 \shift{(1,2)}\der{E}_3 \]
Then $\der{D}_3$ and $\der{E}_3$ are abstraction equivalent. 
\end{restatable}

@@

@@

\begin{definition} 
	Let $\X$ be an  $\mathcal{M}$-adhesive category, the pair $(\X, \mathcal{M})$ is of \emph{type $\mathbb{B}$} if:
	\begin{enumerate}
		\item every arrow in $\mathcal{M}$ is a coproduct coprojection;
		\item $\X$ has all pushouts;
		\item $\X$ has strict initial objects and, for every object $X$, the unique arrow $?_X\colon 0\to X $ belongs to $\mathcal{M}$;
		\item all pushouts are $\mathcal{M}$-stable.
	\end{enumerate}
\end{definition}

@@

\begin{proof}   
	Since $(\X, \R)$ is well-switching, then, there exist, unique, independence pair 
$(i_0,i_1)$ between $\dder{D}_0$ and $\dder{D}_1$,	$(a_0,a_1)$ between $\dder{D}_1$ and $\dder{D}_2$ and	$(e_0, e_1)$ one between $\dder{D}_0$ and
	$S_{a_0,a_1}(\dder{D}_2)$. 	Using \Cref{rem:fill} we can build the diagrams in \Cref{fi:first,fi:second,fi:third}.  The first depicts the  switching $\dder{D}_0$ and $\dder{D}_1$ along  $(i_0, i_1)$, the second that of $\dder{D}_1$ and $\dder{D}_2$ along $a_0$ and $a_1$, while the third refers to $\dder{D}_0$ and
	$S_{a_0,a_1}(\dder{D}_2)$.

@@

		\item In addition to the three diagram above, we have a fourth one, depicted in \Cref{fi:fourth}.
		\begin{figure}[h]
			\centering
			$\xymatrix@C=15pt@R=15pt{&&R_0 \ar@/_.8cm/[ddrr]_(.2){\alpha_0}|(.7)\hole \ar[dr]^{\hspace{-1pt}g_1\circ i_0}&& L_2\ar@/^.8cm/[ddll]^(.2){\alpha_1}  \ar[dl]_{m_2\hspace{-2pt}}\\
				&K_0\ar[dr]^{\hspace{-4pt}q_1\circ v_0}\ar@{>->}[dl]_{l_0} \ar@/_.6cm/[ddrr]|(.36)\hole^(.65){\phi_0}\ar[ur]^{r_0}&& G_2 && K_2 \ar@/^.6cm/[ddll]|(.36)\hole_(.65){\phi_2}\ar[dl]_{k_2}\ar@{>->}[lu]_{l_2} \ar[dr]^{r_2}\\
				L_0 \ar@/_.6cm/[ddrr]_(.2){\gamma_0}|(.31)\hole|(.81)\hole \ar[dr]^(.4){\hspace{-4pt}t_0\circ j_0}|(.6)\hole && Q_1 \ar@{>->}[dl]|(.4)\hole_(.65){s_1}\ar[ur]|(.5)\hole^(.7){t_1}&&D_2 \ar[dr]|(.4)\hole^(.65){g_2} \ar@{>->}[ul]|(.5)\hole_(.65){f_2}&&R_2\ar@/^.6cm/[ddll]^(.2){\gamma_1}|(.31)\hole|(.81)\hole\ar[dl]_(.35){h_2}|(.6)\hole\\&H'_1 &&P_4\ar[dr]_{\lambda_1} \ar@{>->}[dl]^{\lambda_0}\ar[ur]^(.4){\zeta_1}\ar@{>->}[ul]_(.4){\zeta_0}&&G_3\\
				L_2 \ar@/^.6cm/[uurr]^(.2){\alpha_1} \ar[ur]_(.35){\hspace{-2pt}s_1\circ
					\alpha_1}|(.61)\hole&&Q_6
				\ar@{>->}[ul]|(.4)\hole_(.6){\xi_0}\ar[dr]|(.5)\hole^(.65){\xi'_0}
				&&Q_7\ar[ur]|(.4)\hole^(.6){\xi'_1}
				\ar@{>->}[dl]|(.5)\hole_(.65){\xi_1} && R_0 \ar[ul]^(.35){g_2\circ
					\alpha_0}|(.61)\hole\ar@/_.6cm/[uull]_(.2){\alpha_0}\\&K_2
				\ar[ur]_{\lambda_0\circ \phi_2} \ar[dr]_{r_2}
				\ar@{>->}[ul]^{l_2}\ar@/^.6cm/[uurr]_(.65){\phi_2}&&H'_2&& K_0
				\ar[ur]_{r_0} \ar[ul]^{\lambda_1\circ \phi_0} \ar@{>->}[dl]^{l_0}
				\ar@/_.6cm/[uull]^(.65){\phi_0}\\&& R_2
				\ar[ur]_{\hspace{-5pt}\xi_1\circ
					\gamma_1}\ar@/^.8cm/[uurr]^(.2){\gamma_1}&& L_0
				\ar[ul]^{\xi'_0\circ \gamma_0\hspace{-7pt}}
				\ar@/_.8cm/[uull]_(.2){\gamma_0} |(.69)\hole 
			}	$
			\caption{Picture for the proof of \Cref{lem:indep-global-left}: switch of $S_{i_0, i_1}(\dder{D}_0)$ and $\dder{D}_2$.}
			\label{fi:fourth}
		\end{figure}
